Assessing the Replicability of Psychological Science Through Pedagogy

### 2016-08-22 paper/paper.Rmd
spell check

Derermining appropriate sample sizes is a second major challenge in replication research [@button2013;@simonsohn2015]. Post-hoc power analyses are problematic because of the likelihood of inflation of effect sizes due to the "winner's curse" [@hoenig2001;@button2013], but -- especially with the constraints of a limited budget -- it may be impossible to follow more conservative guidelines in all cases. For example, @simonsohn2015 recommends 2.5x the original sample, which can be feasible for small or underpowered original studies but may lead to impractical or unnecessary recommendations for experiments that were initially large and/or adequately powered.

@@

In two cases, in final review of the projects (after data collection), the instructor team decided that the student's choice of key statisical test was not in fact a strong test of the original authors' primary theoretical claim. In both of these cases [@sofer2015;@proudfoot2015], the original authors had fit a multiple regression model to the data and there was not a single obvious test that corresponded directly to the authors' hypothesis. After discussion, the instructor team converged on a statistical test that they thought better corresponded to the original authors' intended hypothesis of interest. We report results from these corrected tests here. We return to this issue in the Discussion, as we believe that test selection is a critical theoretical issue in replication research.

@@

The distribution of ratings are given in Table \ref{tab:results}, along with breakdowns by various factors of interest. Since our sample only includes 11 studies, we describe general trends here rather than attempting to conduct statistical tests (which would be dramatically underpowered after correcting for multiple comparisons). Overall, projects had a somewhat larger probability of being judged successful if it was judged to be closer to the original:
`r 100 * round(mean(d$replicated_TA[d$high_fidelity]), digits=2)`\% 
vs. 
`r 100 * round(mean(d$replicated_TA[d$high_fidelity == FALSE]), digits=2)`\%.
There were small numerical effects of being run on Mechanical Turk (
`r 100 *round(mean(d$replicated_TA[d$original_on_turk == "Yes"]), digits=2)`\%
vs. 
`r 100 *round(mean(d$replicated_TA[d$original_on_turk == "No"]), digits=2)`\%. 
) and having open materials (
`r 100 *round(mean(d$replicated_TA[d$open_materials!="No"]), digits=2)`\% 
vs. 
`r 100 *round(mean(d$replicated_TA[d$open_materials=="No"]), digits=2)`\%), though we do not believe these differences are interpretable given our sample size.

@@

Many statistical factors leading to lowered replicability have been discussed in past work, including analytic flexibility, publication bias, and low statistical power among others [e.g., @button2013,@ioannidis2005]. On the other hand, an alternative viewpoint is that -- especially for studies that leverage social constructs -- exact replication may be difficult across contexts [@gilbert2016;@vanbavel2016]. For both reasons of design and scale, our study here cannot disentangle these factors empirically. Nevertheless, the consistent descreases in effect size and evidential value we report seem consistent with a "winner's curse" -- that intial publications tend to over-estimate effects. And although our sample was too small for statistical inference, numerically our results do not support replication fidelity in terms of sample and procedure as a primary driver of success.

@@

In our view, two recommendations will improve this situation dramatically. First, pre-registration typically requires the selection of one or a small number of statistical analyses to be designated as critical for interpretation. Were this information available, it would have simplified the problem of selecting effects to replicate in several cases. Second, the widespread adoption of open data practices can dramatically facilitate the type of meta-research pursued here, in a number of ways. In addition to clarifying the (often terse) reporting in a paper, the availability of data facilitates the computation of the relevant meta-analytic variables.^[We note that, although we were grateful that several papers in our sample shared data and materials, the use of open (non-proprietary) formats for archving data and analytic code is often critical for other investigators attempting to reproduce analyses.] No paper can perfectly predict the uses to which future investigators will put a piece of work; the availability of the source materials is a way of allowing more flexible reuse.

@@

The goal of this paper was to examine whether people are accurately calibrated in their estimates of how others will respond to their attempts at self-promotion. Our target was Experiment 3, which showed that participants, who were instructed to describe themself in such a way that others would like them, expected to be liked more -- but were actually liked less -- than participants who were asked to describe themselves but not additionally instructed to maximize others’ interest in meeting them. The authors performed a series of regressions on four dependent variables capturing judgements about profile writers (interest, liking, bragging, and success), analyzing these as a function of condition (control vs. "maximize interest") and who the evaluator was (the writer vs. an independent sample). Power analysis for this design was quite complex because the procedure had two stages: in a first stage participants generated descriptions and predicted judgments, while in a second stage a separate sample rated the descriptions.

@@

This paper tested the theory that persistent object representations could assist in spatial navigation, using a navigation paradigm in which participants used keypresses to move though a grid of pictures that were visible only one at a time. In Experiment 1, our target, found that participants located targets faster when navigation involved persistence cues (via sliding animations) than when persistence was disrupted (via temporally matched fading animations). The key test for our replication was a simple *t*-test comparing speed to find the target in the two conditions (slide vs. fade), not controlling for learning across the duration of the experiment. We failed to find this overall difference. However, in an exploratory followup analysis using a linear mixed effect model, we did find a highly-significant effect of condition when controlling for learning epoch. We speculate that the very large effect found by the original authors may have led them to select a simpler analysis that did not control for learning effects. Our data -- which were more variable -- required this control.

@@

This paper reported that the relationship between math anxiety and math performance follows an inverted-U pattern among college undergraduate students with higher intrinsic math motivation, whereas this same relationship is linearly negative among students with lower intrinsic math motivation. We targeted Study 2, the authors' own replication study, in which participants filled out three surveys about their math motivation and anxiet and completed a math task to measure performance. The key effect we targeted was the interaction of math anxiety squared and math motivation in predicting performance. We failed to find evidence for this effect. In contrast to the original study, we found a simple linear trend such that performance decreased with increasing math anxiety for both high and low math-motivation groups.

@@

This paper tested the hypothesis that face typicality affects trustworthiness judgements such that the most typical face will be judged to be the most trustworth. In Study 1, our target, faces of varying typicality were rated on their trustworthiness and attractiveness; while trustworthiness peaked at the most typical face, less typical faces were judged more attractive. While the original class projects selected a global, item-wise regression model as the key statistical test of interest, the instructor team (post-data collection) determined that the fit of this model did not in fact correspond most closely with the authors' stated hypothesis. We thus selected the interaction between face typicality and rating condition as our key test statistic and used the more standard by-participant regression model (which the original paper also reported subsequent to the by-items model) as the target model. Our replication study successfully showed this same interaction. We note that, despite the significant interaction, in our study trustworthiness judgments did not change substantially with typicality (though attractiveness did) and trustworthiness did not peak numerically at the most typical face. Thus, despite replicating one key statistical test, our results are at odds with some of the original papers' claims.

@@

This paper investigated how perceived knowledge affects "overclaiming" (claiming knowledge that you do not have. We targeted Study 1b, which asked participants to rate their knowledge about finances, and then asked them to rate their knowledge about a set of financial terms, some of which were invented (and hence for which a positive knowledge statement would necessarily be an overclaim). The original study found that participants with higher self-rated financial knowledge also overclaimed at a higher rate. The key test statistic for this study was the coefficient in a regression model predicting overclaiming as a function of claimed personal financial knowledge, controlling for accuracy in the true finacial knowledge questions. We successfully replicated this effect, despite our use of a much smaller sample (due to the extremely high post-hoc power of the original study).

@@

This paper investigated whether encouraging participants to think about their personal legacy could increase concern for the environment. The main study of the paper investigated whether a legacy priming writing exercise would increase a variety of environmental measures, including donations to an environmental charity, proenvironmental intentions, and climate change beliefs. The key test we selected was the effect of condition (priming vs. control) on behavioral intentions in an analysis of variance. We observed a marginally significant effect in the same direction in our replication, as well as support for the mediating relationship of legacy motives on beahvioral intentions. Our study may have been underpowered due to budgetary constraints; we ran a sample of comparable size to the original sample of 312 participants. We also may have observed a smaller effect size due to the relatively smaller amount of time our participants spent on the legacy prime writing exercise.

@@

This paper explored constraints on visual working memory in the context of mental rotation. In Experiment 1a, participants perfomed a verbal suppression task to control for use of verbal encoding while attempting to remember a cross with four colored bars, which sometimes switched colors; performing mental rotation significantly impaired ability to detect these bar swaps. To port the task to an online framework we decreased the number of trials and increased the number of participants we tested. The key test statistic was a comparison of K (a metric of the number of visual features remembered) between the rotation and no-rotation conditions. We observed a marginally-significant effect in our replication study. Our study showed a floor effect on capacity such that participants remembered on average very little; it is possible that small details in the displays (or the differing sample available online) led to this floor effect.

>>>>

Determining appropriate sample sizes is a second major challenge in replication research [@button2013;@simonsohn2015]. Post-hoc power analyses are problematic because of the likelihood of inflation of effect sizes due to the "winner's curse" [@hoenig2001;@button2013], but -- especially with the constraints of a limited budget -- it may be impossible to follow more conservative guidelines in all cases. For example, @simonsohn2015 recommends 2.5x the original sample, which can be feasible for small or under-powered original studies but may lead to impractical or unnecessary recommendations for experiments that were initially large and/or adequately powered.

@@

In two cases, in final review of the projects (after data collection), the instructor team decided that the student's choice of key statistical test was not in fact a strong test of the original authors' primary theoretical claim. In both of these cases [@sofer2015;@proudfoot2015], the original authors had fit a multiple regression model to the data and there was not a single obvious test that corresponded directly to the authors' hypothesis. After discussion, the instructor team converged on a statistical test that they thought better corresponded to the original authors' intended hypothesis of interest. We report results from these corrected tests here. We return to this issue in the Discussion, as we believe that test selection is a critical theoretical issue in replication research.

@@

The distribution of ratings are given in Table \ref{tab:results}, along with breakdowns by various factors of interest. Since our sample only includes 11 studies, we describe general trends here rather than attempting to conduct statistical tests (which would be dramatically under-powered after correcting for multiple comparisons). Overall, projects had a somewhat larger probability of being judged successful if it was judged to be closer to the original:
`r 100 * round(mean(d$replicated_TA[d$high_fidelity]), digits=2)`\% 
vs. 
`r 100 * round(mean(d$replicated_TA[d$high_fidelity == FALSE]), digits=2)`\%.
There were small numerical effects of being run on Mechanical Turk (
`r 100 *round(mean(d$replicated_TA[d$original_on_turk == "Yes"]), digits=2)`\%
vs. 
`r 100 *round(mean(d$replicated_TA[d$original_on_turk == "No"]), digits=2)`\%. 
) and having open materials (
`r 100 *round(mean(d$replicated_TA[d$open_materials!="No"]), digits=2)`\% 
vs. 
`r 100 *round(mean(d$replicated_TA[d$open_materials=="No"]), digits=2)`\%), though we do not believe these differences are interpretable given our sample size.

@@

Many statistical factors leading to lowered replicability have been discussed in past work, including analytic flexibility, publication bias, and low statistical power among others [e.g., @button2013,@ioannidis2005]. On the other hand, an alternative viewpoint is that -- especially for studies that leverage social constructs -- exact replication may be difficult across contexts [@gilbert2016;@vanbavel2016]. For both reasons of design and scale, our study here cannot disentangle these factors empirically. Nevertheless, the consistent decreases in effect size and evidential value we report seem consistent with a "winner's curse" -- that initial publications tend to over-estimate effects. And although our sample was too small for statistical inference, numerically our results do not support replication fidelity in terms of sample and procedure as a primary driver of success.

@@

In our view, two recommendations will improve this situation dramatically. First, pre-registration typically requires the selection of one or a small number of statistical analyses to be designated as critical for interpretation. Were this information available, it would have simplified the problem of selecting effects to replicate in several cases. Second, the widespread adoption of open data practices can dramatically facilitate the type of meta-research pursued here, in a number of ways. In addition to clarifying the (often terse) reporting in a paper, the availability of data facilitates the computation of the relevant meta-analytic variables.^[We note that, although we were grateful that several papers in our sample shared data and materials, the use of open (non-proprietary) formats for archiving data and analytic code is often critical for other investigators attempting to reproduce analyses.] No paper can perfectly predict the uses to which future investigators will put a piece of work; the availability of the source materials is a way of allowing more flexible reuse.

@@

The goal of this paper was to examine whether people are accurately calibrated in their estimates of how others will respond to their attempts at self-promotion. Our target was Experiment 3, which showed that participants, who were instructed to describe themselves in such a way that others would like them, expected to be liked more -- but were actually liked less -- than participants who were asked to describe themselves but not additionally instructed to maximize others’ interest in meeting them. The authors performed a series of regressions on four dependent variables capturing judgments about profile writers (interest, liking, bragging, and success), analyzing these as a function of condition (control vs. "maximize interest") and who the evaluator was (the writer vs. an independent sample). Power analysis for this design was quite complex because the procedure had two stages: in a first stage participants generated descriptions and predicted judgments, while in a second stage a separate sample rated the descriptions.

@@

This paper tested the theory that persistent object representations could assist in spatial navigation, using a navigation paradigm in which participants used key presses to move though a grid of pictures that were visible only one at a time. In Experiment 1, our target, found that participants located targets faster when navigation involved persistence cues (via sliding animations) than when persistence was disrupted (via temporally matched fading animations). The key test for our replication was a simple *t*-test comparing speed to find the target in the two conditions (slide vs. fade), not controlling for learning across the duration of the experiment. We failed to find this overall difference. However, in an exploratory followup analysis using a linear mixed effect model, we did find a highly-significant effect of condition when controlling for learning epoch. We speculate that the very large effect found by the original authors may have led them to select a simpler analysis that did not control for learning effects. Our data -- which were more variable -- required this control.

@@

This paper reported that the relationship between math anxiety and math performance follows an inverted-U pattern among college undergraduate students with higher intrinsic math motivation, whereas this same relationship is linearly negative among students with lower intrinsic math motivation. We targeted Study 2, the authors' own replication study, in which participants filled out three surveys about their math motivation and anxiety and completed a math task to measure performance. The key effect we targeted was the interaction of math anxiety squared and math motivation in predicting performance. We failed to find evidence for this effect. In contrast to the original study, we found a simple linear trend such that performance decreased with increasing math anxiety for both high and low math-motivation groups.

@@

This paper tested the hypothesis that face typicality affects trustworthiness judgments such that the most typical face will be judged to be the most trustworthy. In Study 1, our target, faces of varying typicality were rated on their trustworthiness and attractiveness; while trustworthiness peaked at the most typical face, less typical faces were judged more attractive. While the original class projects selected a global, item-wise regression model as the key statistical test of interest, the instructor team (post-data collection) determined that the fit of this model did not in fact correspond most closely with the authors' stated hypothesis. We thus selected the interaction between face typicality and rating condition as our key test statistic and used the more standard by-participant regression model (which the original paper also reported subsequent to the by-items model) as the target model. Our replication study successfully showed this same interaction. We note that, despite the significant interaction, in our study trustworthiness judgments did not change substantially with typicality (though attractiveness did) and trustworthiness did not peak numerically at the most typical face. Thus, despite replicating one key statistical test, our results are at odds with some of the original papers' claims.

@@

This paper investigated how perceived knowledge affects "overclaiming" (claiming knowledge that you do not have. We targeted Study 1b, which asked participants to rate their knowledge about finances, and then asked them to rate their knowledge about a set of financial terms, some of which were invented (and hence for which a positive knowledge statement would necessarily be an overclaim). The original study found that participants with higher self-rated financial knowledge also overclaimed at a higher rate. The key test statistic for this study was the coefficient in a regression model predicting overclaiming as a function of claimed personal financial knowledge, controlling for accuracy in the true financial knowledge questions. We successfully replicated this effect, despite our use of a much smaller sample (due to the extremely high post-hoc power of the original study).

@@

This paper investigated whether encouraging participants to think about their personal legacy could increase concern for the environment. The main study of the paper investigated whether a legacy priming writing exercise would increase a variety of environmental measures, including donations to an environmental charity, pro-environmental intentions, and climate change beliefs. The key test we selected was the effect of condition (priming vs. control) on behavioral intentions in an analysis of variance. We observed a marginally significant effect in the same direction in our replication, as well as support for the mediating relationship of legacy motives on behavioral intentions. Our study may have been under-powered due to budgetary constraints; we ran a sample of comparable size to the original sample of 312 participants. We also may have observed a smaller effect size due to the relatively smaller amount of time our participants spent on the legacy prime writing exercise.

@@

This paper explored constraints on visual working memory in the context of mental rotation. In Experiment 1a, participants performed a verbal suppression task to control for use of verbal encoding while attempting to remember a cross with four colored bars, which sometimes switched colors; performing mental rotation significantly impaired ability to detect these bar swaps. To port the task to an online framework we decreased the number of trials and increased the number of participants we tested. The key test statistic was a comparison of K (a metric of the number of visual features remembered) between the rotation and no-rotation conditions. We observed a marginally-significant effect in our replication study. Our study showed a floor effect on capacity such that participants remembered on average very little; it is possible that small details in the displays (or the differing sample available online) led to this floor effect.
